Data Specialization for a Staged Lambda Calculus

### 2015-02-21 paper/intro.tex
partway through editing section 2
culling some old intro text

It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).  This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).  Examples include:
\begin{itemize}
\item Computing $b^e$, where $b$ varies and $e$ is constant.
\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
\end{itemize}

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

Once we have appropriately written {\em multi-stage} programs, we then must execute them in a way that respects the staging.  In this paper, we explore two techniques for this: {\em partial evaluation} and {\em stage-splitting}.

The main idea of the former technique is to evaluate all of the constant portions of the program and embed the partial results into a {\em residual} program that depends only on the varying inputs.  This is an old and well-explored idea, though the typical setup of partial evaluation implies a more powerful binding time analysis than is necessary for our language.

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.  This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruf\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

\TODO motivate this example

\TODO integrate the following text into the new intro

@@

\ur{  Based on this definition, it
  might be good and correct to argue that this is more or less an
  orthogonal problem.}

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before {\em any arguments} are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

>>>>

\crem{Notes from 2/21:
\begin{enumerate}
\item see K/R intro.
\item function specialization is an important optimization technique which can
be performed whether or not the function's arguments are known.
\item data specialization statically specializes a multi-argument program
\emph{before} having any of its arguments, unlike partial evaluation.
\item we are doing data specialization, but on lambda calculus (cf C-like
languages, as in graphics papers).
\item this additional expressive power gives us some interesting
results---asymptotic speedup on some algorithms (and generation of recursive
types from recursive functions), and cross-cutting (?) modularity (see graphics
applications).
\end{enumerate}
}

\subsection{Old intro}

partial evaluation requires compile-time access to first stage, whereas we don't

Prior work on metaprogramming allows us to evaluate such staged code
in a staged fashion by first providing the first-stage arguments and
generating a piece of code that is specialized for the arguments.  For
example, we can evaluate the staged quickselect by providing a
list~\texttt{l} to obtain a piece of code that then can be used with
different ranks (\texttt{k} values) in the second stage.
Such a specialized code essentially performs all the first-stage computations
first and uses their results to improve the efficiency in the second stage,
where we issue many queries.

For example, in our quickselect example, it is unlikely to know the
list~\texttt{l} that we will be operation on at compile time.

\cite{knoblock96} proposed the {\em data specialization}
technique to overcome some of these limitations of program
specialization.  
Unlike program specialization,
data specialization does not require knowledge of the first-stage
arguments to produce fast code for the second stage.  
Instead it yield
a first stage program that maps the first stage arguments to an
intermediate or {\em boundary data structure}, which is used by the
second stage program along with the second stage arguments to produce
the final results.

But their work applies to program fragments that don't involve sophisticated
data types such as as sums and recursive types, and don't involve recursion.

\subsection{Oldest intro}

This most often arises when
we want to evaluate a multivariate function several times, with some inputs
dynamic (varying across invocations) and some static (constant across
invocations).
\begin{itemize}
\item Computing $b^e$, where $b$ varies and $e$ is constant.
\item Finding the $i$th largest element in a list, where the list is constant
and $i$ varies.
\item Computing the shading of a surface in a 3D rendering, where some
contributors to the final pixel color can be approximated as constant over
pixel blocks.  
\end{itemize}
Prior examples
include the ``data specialization" work of Knoblock and Ruf~\cite{knoblock96}
as well as the compilation of high-level shading languages for computer graphics
such as RTSL~\cite{Proudfoot:2001} and Spark~\cite{Foley:2011}.

@@

Taking a multivariate function $f$ as input, stage-splitting produces two
functions $f_1$ and $f_2$, where $f_1$ uses the stage-\bbone\ input to produce a
data structure, and $f_2$ consumes that data structure and the stage-\bbtwo\
input to produce the final output.  More precisely, a stage splitter is any $s$
such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
The stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.
